**The statistical analysis of non-linear longitudinal data in biomedical research using generalized additive models**

### 2021-05-27 Manuscripts/Manuscript by chapters -v2/01-Background.Rmd
Making proposed changes in the manuscript

The expected linear behavior of the response through time is a key requisite in rm-ANOVA [@pinheiro2006]. This "linearity assumption" in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference. In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies. A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009]. These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation. Therefore, when rm-ANOVA is used to draw inference of such highly-variable data the estimates are inevitably biased, because the model is only able to accommodate linear trends that are far from adequately representing the biological phenomenon of interest.

A _post hoc_ analysis is the statistical test used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a _p-value_, which in turn is used as a measure of significance.
Although it is possible that a _post hoc_ analysis of rm-ANOVA is able to find “significant” _p-values_( _p_<0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data. In other words, _p-values_ are valid only if the model and the data have good agreement; if that is not the case, a "Type III" error (known as "model misspecification") occurs[@dennis2019]. For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting _p-values_ and parameter estimates to be invalid [@wang2019].

@@

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal information. The recognition on the shortcomings of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data [@skala2010;@vishwanath2009]. Briefly, LMEMs incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age). When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006]. However, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013]. And even more importantly, LMEMs also expect a linear relationship between the response and time [@pinheiro2006], making them unsuitable to analyze non-linear data.

>>>>

The expected linear behavior of the response through time is a key requisite in rm-ANOVA [@pinheiro2006]. This "linearity assumption" in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference. In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies. A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009]. These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation. Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.

A _post hoc_ analysis is often used in conjunction with rm-ANOVA to perform repeated comparisons to estimate a _p-value_, which in turn is used as a measure of significance.
Although it is possible that a _post hoc_ analysis of rm-ANOVA is able to find “significant” _p-values_( _p_<0.05) from non-linear data, the validity of such metric is dependent on how adequate the model fits the data. In other words, _p-values_ are valid only if the model and the data have good agreement; if that is not the case, a "Type III" error (known as "model misspecification") occurs[@dennis2019]. For example, model misspecification will occur when a model that is only able to explain linear responses (such as rm-ANOVA) is fitted to data that follows a quadratic trend, thereby causing the resulting _p-values_ and parameter estimates to be invalid [@wang2019].

@@

During the last decade, the biomedical community has started to recognize the limitations of rm-ANOVA in the analysis of longitudinal data. The recognition on the shortcomings of rm-ANOVA is exemplified by the  use of  linear mixed effects models (LMEMs) by certain groups to analyze longitudinal tumor response data [@skala2010;@vishwanath2009]. Briefly, LMEMs incorporate _fixed effects_, which correspond to the levels of experimental factors in the study (e.g., the different drug regimens in a clinical trial), and _random effects_, which account for random variation within the population (e.g., the individual-level differences not due to treatment such as weight or age). When compared to the traditional rm-ANOVA, LMEMs are more flexible as they can accommodate missing observations for multiple subjects and allow different modeling strategies for the variability within each measure in every subject [@pinheiro2006]. However, LMEMs impose restrictions in the distribution of the errors  of the random effects, which need to be normally distributed and independent [@gueorguieva2004;@barr2013]. And even more importantly, LMEMs also assume a linear relationship between the response and time [@pinheiro2006], making them unsuitable to analyze non-linear data.

### 2021-05-27 Manuscripts/Manuscript by chapters -v2/02-Challenges.Rmd
Making proposed changes in the manuscript

In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively. Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ effects, and are assumed to be $\sim N(0,\sigma^2)$ (independently normally distributed with mean zero and variance $\sigma^2_\mu$).
In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs. "novel drug" or "saline" vs. "chemotherapy"). Then, the group terms in Equation \@ref(eq:linear-model) can be written as below with $treatment_j=0$ representing the first treatment group (Group A) and $treatment_j=1$ representing the second treatment group (Group B). The linear models then can be expressed as

@@

Missing observations are an issue that arises frequently in longitudinal studies. In biomedical research, this situation can be caused by reasons beyond the control of the investigator [@molenberghs2004]. Dropout from patients and attrition or injury in animals are among the reasons for missing observations. Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].  In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest [@scheffer2002]. If the data are MCAR, this means that the missingness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value. An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis. This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].

In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR [@west2014]. For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because  the missigness is unrelated to other variables of interest.

This section has presented the assumptions for analyzing longitudinal data using rm-ANOVA and LMEMs and compared their differences regarding linearity, the covariance matrix and  missing data. In particular, LMEMs offer a more robust and flexible approach than rm-ANOVA and if the data follows a linear trend, they provide an excellent choice to derive inferences from a repeated measures study. However, when the data presents high a non-linear behavior, LMEMs and rm-ANOVA fail to capture the trend of the data. To better convey the issues of linearity and correlation  in linear models fitted to non-linear data, simulation is used in the next section.

@@

To demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each. An rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme]. 
Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response), A). In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:l-q-response), C). In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point. This is intentional in order to simulate the expected temporal evolution of some physiological quantity.

Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as therapy progresses a divergence in the trend of the response indicates a treatment effect. In other words, Group 1 can be thought as a "Control" group and Group 2 as a "Treatment" group. From the mean response per group (linear or quadratic), the variability or "error" of individual responses  within each group is simulated using a covariance matrix with compound symmetry (constant variance across time). Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \@ref(fig:l-q-response) B,D). A more comprehensive exploration  of the fit of rm-ANOVA for linear and non-linear longitudinal appears in Figure \@ref(fig:linear-cases-Appendix) and Figure \@ref(fig:quadratic-cases-Appendix) in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented.

@@

A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure ((\@ref(fig:l-q-response), D) indicates that the model is not capturing the changes within each group in a good way.  Specifically, note that the fitted mean response of the rm-ANOVA model (panel D) shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model. Moreover, the rm-ANOVA model is not being able to capture the fact that the initial values are the same in each group, and instead fits non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panels C and D). If such change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response. Thus, even though the model correctly detects a difference in treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.

>>>>

In this model $y_{ijt}$ is the response for subject $i$, in treatment group $j$ at time $t$, which can be decomposed in a mean value $\beta_0$, _fixed effects_ of time ($time_t$), treatment ($treatment_j$) and their interaction $time_t*treatment_j$ which have linear slopes given by $\beta_1, \beta_2$ and $\beta_3$, respectively. Independent errors $\varepsilon_{tij}$ represent random variation not explained by the _fixed_ effects, and are assumed to be $\sim N(0,\sigma^2)$ (independently and identically normally distributed with mean zero and variance $\sigma^2$).
In a  biomedical research context, suppose two treatments groups are used in a study (e.g., "placebo" vs. "novel drug" or "saline" vs. "chemotherapy"). Then, the group terms in Equation \@ref(eq:linear-model) can be written as below with $treatment_j=0$ representing the first treatment group (Group A) and $treatment_j=1$ representing the second treatment group (Group B). The linear models then can be expressed as

@@

Missing observations are an issue that arises frequently in longitudinal studies. In biomedical research, this situation can be caused by reasons beyond the control of the investigator [@molenberghs2004]. Dropout from patients and attrition or injury in animals are among the reasons for missing observations. Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].  In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest [@scheffer2002]. If the data are MCAR, this means that the missingness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value.

An rm-ANOVA model assumes complete observations for all subjects, and therefore subjects with one or more missing observations are excluded from the analysis. This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012]. In the case of LMEMs, inferences from the model are valid when missing observations in the data exist that are MAR or MCAR [@west2014]. For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because  the missigness is unrelated to other variables of interest.

@@

To visually demonstrate the limitations of rm-ANOVA an LMEMs for non-linear longitudinal data, this section presents a simulation experiment of a normally distributed response of two groups of 10 subjects each. An rm-ANOVA model (Equation \@ref(eq:linear-model)) is fitted to each group, using $\textsf{R}$[@r] and the package _nlme_[@nlme]. 
Briefly, two cases for the mean responses for each group are considered: in the first case, the mean response in each group is a linear function with different intercepts and slopes; a negative slope is used for Group 1 and a positive slope is used for Group 2 (Figure \@ref(fig:l-q-response), A). In the second case, a second-degree polynomial (quadratic) function is used for the mean response per group: the quadratic function is concave down for Group 1 and it is concave up for Group 2 (Figure \@ref(fig:l-q-response), C). In both the linear and quadratic simulated data, the groups start with the same mean value at the first time point. This is intentional in order to simulate the expected temporal evolution of some physiological quantity, which is typical in biomedical experiments.

Specifically, the rationale for the chosen linear and quadratic functions is the likelihood that a measured response in two treatment groups is similar in the initial phase of the study, but as therapy progresses a divergence in the trend of the response indicates a treatment effect. In other words, Group 1 can be thought as a "Control" group and Group 2 as a "Treatment" group. From the mean response per group (linear or quadratic), the variability or "error" of individual responses  within each group is simulated using a covariance matrix with compound symmetry (constant variance across time). Thus, the response per subject in both the linear and quadratic simulation corresponds to the mean response per group plus the error (Figure \@ref(fig:l-q-response) B,D).

A more comprehensive exploration  of the fit of rm-ANOVA for linear and non-linear longitudinal appears in Figure \@ref(fig:linear-cases-Appendix) and Figure \@ref(fig:quadratic-cases-Appendix) in the Appendix, where simulation with compound symmetry and independent errors (errors generated from a normal distribution that are not constant over time) and the plot of simulated errors, and fitted parameters in presented. We are aware that the simulated data used in this section present an extreme case that might not occur frequently in biomedical research, but they are used as a representation of the consequences of modeling non-linear data with a linear model such as rm-ANOVA or LMEMs. Of notice, Section \@ref(longitudinal-GAMs) uses simulated data that does follow reported trends in the biomedical literature.

@@

A comparison of the fitted mean response of the rm-ANOVA model to the the simulated data in Figure ((\@ref(fig:l-q-response), D) indicates that the model is not capturing the changes within each group.  Specifically, note that the fitted mean response of the rm-ANOVA model (panel D) shows that the change (increase for Treatment 1 or decrease for Treatment 2) in the response through time points 2 and 4 is not being captured by the model. Moreover, the rm-ANOVA model is not being able to capture the fact that the initial values are the same in each group, and instead fits non-parallel lines that have initial values that are markedly different from the "true" initial values in each case (compare panels C and D). If such a change has important physiological implications,the rm-ANOVA model omits it from the fitted mean response. Thus, even though the model correctly detects a difference in treatment groups, the exact nature of this difference is not correctly identified, limiting valuable inferences from the data.

### 2021-05-27 Manuscripts/Manuscript by chapters -v2/03-GAM Theory.Rmd
Making proposed changes in the manuscript

Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the function $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the residual error.

In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible _smooth functions_. This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response. One possible set of functions for $f(x_t\mid \beta_j)$ that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998]. In particular, polynomial fits are known to show boundary effects because as $t$ goes to $\pm \infty$, $f(x_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic, and causes bias at the endpoints of the time period.

The smooth functional relationship between the covariates and the response in GAMs is specified   using a semi-parametric relationship that can be fit within the GLM framework, by using _basis functions_ expansions of the covariates and by estimating random coefficients for these basis functions. A _basis_ is a set of functions that spans the space where the smooths that approximate $f(x_t \mid \beta_j)$ exist [@simpson2018]. For the linear model in Equation \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $time_t$, $treatment_j$ and $time_t \times treatment_j$. The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response [@hefley2017], which in the case of Equation \@ref(eq:linear-model) is restricted to a linear family of functions.  In the case of Equation \@ref(eq:GAM), the basis function is $f(x_t\mid \beta_j)$, which means that the model allows for non-linear relationships among the covariates.

>>>>

Where $y_{ijt}$ is the response at time $t$ of subject  $i$ in group $j$, $\beta_0$ is the expected value at time 0, the change of $y_{ijt}$ over time is represented by the _smooth function_ $f(x_t\mid \beta_j)$ with inputs as the covariates $x_t$ and parameters $\beta_j$, and $\varepsilon_{ijt}$ represents the residual error.

In contrast to the linear functions used to model the relationship between the covariates and the response in rm-ANOVA or LMEM, GAMs use more flexible _smooth functions_. This approach is advantageous as it does not restrict the model to a linear relationship, although a GAM will estimate a linear relationship if the data is consistent with a linear response. One possible set of functions for $f(x_t\mid \beta_j)$ that allow for non-linear responses are polynomials, but a major limitation is that polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with inference [@beck1998]. In particular, polynomial fits are known to show boundary effects because as $t$ goes to $\pm \infty$, $f(x_t \mid \beta_j)$ goes to $\pm \infty$ which is almost always unrealistic, and causes bias at the endpoints of the time period.

The smooth functional relationship between the covariates and the response in GAMs is specified   using a semi-parametric relationship that can be fit within the GLM framework, by using _basis function_ expansions of the covariates and by estimating random coefficients associated with these basis functions. A _basis_ is a set of functions that spans the mathematical space where the smooths that approximate $f(x_t \mid \beta_j)$ exist [@simpson2018]. For the linear model in Equation \@ref(eq:linear-model), the basis coefficients are $\beta_1$, $\beta_2$ and $\beta_3$ and the basis vectors are $time_t$, $treatment_j$ and $time_t \times treatment_j$. The basis function then, is the combination of basis coefficients and basis vectors that map the possible relationship between the covariates and the response [@hefley2017], which in the case of Equation \@ref(eq:linear-model) is restricted to a linear family of functions.  In the case of Equation \@ref(eq:GAM), the basis function is $f(x_t\mid \beta_j)$, which means that the model allows for non-linear relationships among the covariates.
